Outline flowCore paper

### 2008-05-07 paper/manuscript.tex
Changed the introduction a little and added a few sentences about plateCore.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31535 bc3139a8-67e5-0310-9ffc-ced21a209358

\section*{Introduction}
Traditionally, flow cytometry (FCM) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry high content screening (FC-HCS)
have recently been developed and are now used in both basic and
clinical research. As a result, the need for flexible and well
structured computational tools to efficiently handle and analyze
FC-HCS has increased considerably.
%Indeed, the amount of information generated by these technologies must be stored, managed, and needs to be summarized in order to be accessible to the researcher. This sentance is not grammaically corect, and really doesn't add anything anyway
However, the absence of a shared research platform that
bioinformaticians, computer scientists, and statisticians can utilize
to develop novel \ref{lizard2007fca}.
% or standard - not sure we want to advocate standard
% methods? Do we really want to say everyone should use the same method
% for analysis?  methods for flow cytometric analysis has hindered
% innovation. <ref Lizard G. FCM analyses and bioinformatics: Interest
% in new softwares to optimize novel technologies and to favor the
% emergence of innovative concepts in cell research. Cytometry A
% 2007;71A:646–647> \ref{lizard2007fca}
We believe that the open source statistical
software R, in conjunction with the Bioconductor Project can fill this
gap.  In this paper, we propose R data structures to handle FCM data
through the main steps of importing, storing, assessing and
preprocessing. These structures, along with a range of specialized
methods for compensation, transformation, gating and additional
preprocessing steps, are implemented in the newly developed
Bioconductor package flowCore.  flowCore is a freely available,
sophisticated and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and enables open
development of new methods and tools for their coherent
analysis. flowCore runs on Windows, Mac OS X and Linux/Unix operating
systems.

@@

\subsection*{Existing data standards and conventions}
Currently, data from FCM experiments are stored in single files
according to the Flow Cytometry Standard (FCS) \ref{seamer1997pnd}.
%This standard attempts to capture much of the accompanying meta information, most of which is produced by the measurement instrument or has to be supplied manually by the operator upon experiment setup. - I believe FCS only tries to capture a subset of meta information
However, the recent
developments in high-throughput FCM are shifting the focus
of interest away from single-tube based measurements towards large and
complex experimental designs with dozens of co-variates and
influencing factors. Self-contained and self-reflective data
structures are a prerequisite to allow for coordinated analysis of
such experiments. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of %most of <can you think of an example that is not closed source?>
these products makes it impractical to
connect to and integrate the data into an analysis pipeline.

@@

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  For example, the Gating-ML Candidate Recommendation (CR)
represents a proposal on how to form unambiguous XML-based gate
definitions, which can facilitate the interchange and validation of
data between different software packages with the potential of
significant increase of hardware and software interoperability. Gates
may be ordered into a hierarchical structure to describe a gating
strategy. They may be applied on parameters as in FCS files or on
transformed parameters as described by an explicit parameter
transformation. 
%% I find the previous sentence redundant - 
%% They may be applied on raw or transformed parameters
This enables the encoding of a simple analytical workflow, in a way
that allows us to reconstruct the analysis programmatically.

@@

\subsection*{Flow Set}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of common errors % (\textit{i.e.}, example of an error ) 
that occur when using more or less ad hoc spreadsheets that
lack a controlled structure. There is no native ordering
to a flow set, which in a way reflects the still predominant
application of tube-based experiments rather than experiments
following a plate-design. However, the flow set structure can be
readily extended to also accommodate such spatial information
\textit{(plateCore reference, Perry, Errol?)}.

>>>>

\section*{Introduction}
Traditionally, flow cytometry (FCM) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry high content screening (FC-HCS)
have recently been developed and are now used in both basic and
clinical research. As a result, the need for flexible and well
structured computational tools to efficiently handle and analyze
FC-HCS has increased considerably.
However, the absence of a shared research platform that
bioinformaticians, computer scientists, and statisticians can collaboratively
use has hindered the development of novel methods 
for FCM data analysis \citep{lizard2007fca}.
We believe that the open source statistical
software R, in conjunction with the Bioconductor Project can fill this
gap.  
%% EAS: I think we mention flowCore before talking about data structures and
% methods.  We should first say explicity that we have a solution for
% the missing shared platform.
We have developed the Bioconductor package flowCore as a freely available,
sophisticated and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. 
%In this paper, we propose R data structures to handle FCM data
%through the main steps of importing, storing, assessing and
%preprocessing. 
%These structures, along with a range of specialized
%methods for compensation, transformation, gating and additional
%preprocessing steps, are implemented in the newly developed
%Bioconductor package flowCore.  flowCore is a freely available,
%sophisticated and extensible FCM data analysis platform that enables
%researchers to efficiently handle FC-HCS data and enables open
%development of new methods and tools for their coherent
%analysis.
flowCore runs on Windows, Mac OS X and Linux/Unix operating
systems.

@@

\subsection*{Existing data standards and conventions}
Currently, data from FCM experiments are stored in single files
according to the Flow Cytometry Standard (FCS) \citep{seamer1997pnd}.
%This standard attempts to capture much of the accompanying meta information, most of which is produced by the measurement instrument or has to be supplied manually by the operator upon experiment setup. - I believe FCS only tries to capture a subset of meta information
However, the recent
developments in high-throughput FCM are shifting the focus
of interest away from single-tube based measurements towards large and
complex experimental designs with dozens of co-variates and
influencing factors. Self-contained and self-reflective data
structures are a prerequisite to allow for coordinated analysis of
such experiments. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of %most of <can you think of an example that is not closed source?>
these products makes it impractical to
connect to and integrate the data into an analysis pipeline.

@@

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  For example, the Gating-ML Candidate Recommendation (CR)
represents a proposal on how to form unambiguous XML-based gate
definitions, which can facilitate the interchange and validation of
data between different software packages with the potential of
significant increase of hardware and software interoperability. Gates
may be ordered into a hierarchical structure to describe a gating
strategy. They may be applied on parameters as in FCS files or on
transformed parameters as described by an explicit parameter
transformation. 
%% I find the previous sentence redundant - 
%% They may be applied on raw or transformed parameters
%% EAS: I don't believe the previous sentence is required.
This enables the encoding of a simple analytical workflow, in a way
that allows us to reconstruct the analysis programmatically.

@@

\subsection*{Flow Set}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of common errors % (\textit{i.e.}, example of an error ) 
that occur when using more or less ad hoc spreadsheets that
lack a controlled structure. There is no native ordering
to a flow set, which in a way reflects the still predominant
application of tube-based experiments rather than experiments
following a plate-design. 
%% EAS: I don't think the previous sentence is required. Does it 
%% matter if the flowSet is ordered or not?

%However, the flow set structure can be
%readily extended to also accommodate such spatial information
%% EAS: The spatial aspect of plateCore is really secondary to having 
%% an easy, consistent way to manage and access the mulitplex annotation.
%% Also, it's important to have a standard format for the annotation if
%% we're trying to encourage collaboration. I haven't done it yet, but I 
%% intend to look into
%% into stealing the annotation implementation from CellHTS2 and
%% using it in plateCore.
However, the flow set structure can be readily extended to incorporate the
potentially complex annotation associated with large FC-HCS experiments. Access
to flowCore data structures and methods has already enabled the development of
new Bioconductor pacakges, such as plateCore, that faciliate the analysis of
plate-based FC-HCS data and make handling the annotation and processing
the data easier.

%\textit{(plateCore reference, Perry, Errol?)}.
